---
title: "Final Project - Module 8"
author: "Rtists"
date: "2022-12-03"
output: html_document
bibliography: BIOL3140.Rtists.finalproj.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE,warning = FALSE)
```

## Introduction
Climate change may be having a detrimental impact on the conservation of the trans-Gulf migrant (TGM) population. TGMs are neotropical passerines that migrate from Central and South America to temperate latitudes during the northern hemisphere spring, flying non-stop over the Gulf of Mexico and arriving on land between Texas and Florida. TGMs make this journey in order to forage and breed in the temperate latitudes. According to @smith2005arrival and @newton2008the, Evolutionary trade-offs are made depending on if the TGM arrives earlier and has an increased chance of finding a mate and breeding multiple times, or later and having an increased food availability. The dates that TGMs arrive at their destination has become of particular interest to researchers because of these consequences on fitness, and the way that climate change is influencing these arrival dates. According to @both2006climate, spring temperatures are rising, and TGM populations may decline if they fail to shift their breeding site arrival dates accordingly.
There is debate over if and how TGMs and other migrants shift their spring arrival in temperate North America in response to local temperature changes. A 2019 study, @horton2019holding, did not find an overall change in the annual number of migrants between 2007 and 2015 or the annual timing of peak migration from 1995 to 2015; however, they did find that the earliest seasonal movements throughout the region occurred a significant 1.6 days per decade earlier. Also, according to @horton2019holding, the study also found phenological changes over time; the most significant being the emergence of larger-bodies and shorter distance migrants. Despite this study, the topic of migration pattern changes in response to climate change is still hotly debated and needs to be researched further. 
In 2002, the Cornell Laboratory of Ornithology and the National Audubon Society launched eBird, the most popular and far-reaching online program through which amateur birders can submit bird observations. This data supplements what is collected through field, banding, and tracking studies, which are often temporally limited due to the small number of trained experts. This data aims to study the effects of changes in climate change linked local environment parameters on migration phenology. The data collected by eBird covers a wide range of taxonomy and geography. Thousands of amateur birders have contributed over 100 million bird observations to eBird, including historical observations that predate the launch of the program.
We downloaded eBird’s species occurrence data from the Global Biodiversity Information Facility with an R package, rgif, that allows us to access the application programming interface. We also used an R package, rnoa, to interact with NOAA’s National Climate Data Center’s application programming interface in order to compile meteorological data. We used this eBird and meteorological data to study the effects of local weather conditions on the arrival times of TGMs in Massachusetts. We were investigating if local weather conditions that are related to climate change are impacting TGM arrival times, which has implications for fitness changes and species conservation.

## Methods
The first step in completing our project was to select five different trans-Gulf migrant (TGM) bird species. These species should be from different families. We selected: Belted Kingfisher (*Megaceryle alcyon*), Eastern Phoebe (*Sayornis phoebe*), Philadelphia Vireo (*Vireo phildelphicus*), Swainson's Thrush (*Catharus ustulatus*), Black-Throated Blue Warbler (*Setophaga caerulescens*) and Yellow-Throated Warbler (*Setophaga dominica*). Data would be sourced from human observations of these species in Massachusetts in the eBird database. Then, using a for loop that passes our species through occ_data(), we retrieved the data from the database. This information would allow us to find the Julian day for each species and see the effect of temperature and wind along the migration route to Massachusetts. Next three sample locations along the migration route are chosen: Mobile, AL, Charlotte, NC, and Boston, MA. Mobile, AL is on the Gulf, Charlotte, NC is roughly in the middle, and Boston, MA is the endpoint. Weather data from each of these locations was downloaded using a token from the NOAA website. Next the eBird data is plotted the proportion of the population that arrived and the predicted arrival time for each. Then another plot is created to illustrate how the Julian day changes over the years. This is then used to see the impact of climate on the arrival date. Weather data was then used to calculate wind direction and results were joined with the species data. The frollmean() function is then used to calculate the mean of the weather variables 5 and 10 days before the arrival day. Lastly, a Linear Mixed-effect Model was created using lme for both weather data sets (single-day and two-week average). An Anova test was run to find the best fit model, and model testing was done using the function dredge().

## Results
```{r,"loading in libraries"}
library(rgbif)
library(tidyverse)
library(MuMIn)
library(rnoaa)
library(data.table)
library(ggmap)
library(usmap)
library(magick)
library(cowplot)
library(lme4) 
library(car)
library(knitr)
```
```{r,"getting specific bird data--did this already in Rscript",eval=FALSE}
#I put this on eval=FALSE because I already loaded everything in an RScript and then saved it as RDS, so I've been using the RDS for this document and didn't want to have to reload all the data in. 
species <- c("Megaceryle alcyon","Sayornis phoebe","Vireo philadelphicus","Catharus ustulatus","Setophaga caerulescens","Setophaga dominica")
y <- paste0("1990",",","2019")
m <- paste0("3",",","6")
dat.l <- list()
for (i in species){
  n.obs <- occ_data(scientificName = i,year = y,month = m,limit=0,country = "US",basisOfRecord = "HUMAN_OBSERVATION", stateProvince = "Massachusetts")$meta$count
  print(n.obs)
  dat.l[[paste0(i)]] <- occ_data(scientificName = i, year=y, month = m, limit = n.obs,country = "US",basisOfRecord = "HUMAN OBSERVATION",stateProvince = "Massachusetts")[[2]]
}
data = rbindlist(dat.l,fill = T)
saveRDS(data,"massbird.data.RDS")
```

```{r,"making data table from data + general plot of how many species v. year",cache=TRUE}
dat<- readRDS("massbird.data.RDS")
dat%>%
  group_by(year,species)%>%
  summarise(count=sum(individualCount,na.rm = T))%>%
  ggplot(aes(x=year,y=count,col=species))+geom_point()
```


After loading in the species occurrence data from GBIF, this graph displays a general overview of the quantity of data points we have. It shows the count many species occurrence recordings there were per year.  

```{r,"loading in weather station data + making into table",cache=TRUE}
options(noaakey="OOZIuxBSmRgSCBNQPuYyWVUZGaWoDIBo")
sts <- c(
  "GHCND:USW00013894", #Mobile, AL 2k away about 10 days away @200 km/day
  "GHCND:USW00013881", #Charlotte, NC 1000 km away about 6 days away @200 km/day
  "GHCND:USW00014739" #Boston
)
sta.d <- bind_rows( #bind the rows
  lapply(sts,function(x) ncdc_stations(stationid = x)$data ) #use lapply to run through stations
)%>%
  mutate(usmap_transform(.,input_names = c("longitude","latitude"),output_names = c("longitude.1", "latitude.1")))%>% #join transformation of lat/long for projection with usmap
  mutate(name=str_sub(name, -5,-4))%>%#simplify the name column, grab just the state
  mutate(migr.day=c(10,5,0))%>% #so we can look at wind speed 0, 5 or 10 days before arrive in boston
  separate(id,into = c("station.type","id"))#need to cut station type out from station id number
```
```{r,"graphing stations onto map"}
plot_usmap(include = c(.northeast_region,.south_region,.east_north_central))+geom_point(data=sta.d,aes(x=longitude.1,y=latitude.1,col=name),size=5)+geom_label(data=sta.d,aes(x=longitude.1,y=latitude.1,col=name,label=name),size=5,nudge_x = 1e6*0.25)+theme(legend.position = "none")

```


This map is a graphical representation of the locations of the 3 stations we are collecting weather data out of. 

```{r,"getting weather data from stations",cache=TRUE}
weather.d <- meteo_pull_monitors(sta.d$id,date_min = "2000-01-01")
```
```{r,"all of the species arrival"}
all.dat<- dat%>%
  group_by(year)%>%
  mutate(date=as.Date(paste0(year,"-",month,"-",day)),
         j.day=julian(date,origin=as.Date(paste0(unique(year),"-01-01")))
  )%>%
  group_by(species,year,j.day,date)%>%
  summarise(day.tot=sum(individualCount,na.rm=T))%>%
  group_by(species,year)%>%
  mutate(prop=cumsum(day.tot/sum(day.tot,na.rm = T)))%>%
  filter(year>1999)
all.dat%>%
  ggplot(aes(j.day,prop,color=species))+geom_point()+facet_wrap(year~.)
```

This is a graphical representation of the arrival of each species over the span of 2000-2019. Below, we divided this arrival data by species and created logistic models to model each species' arrival data and estimate the arrival function. 

```{r, "modeling graphs for each species", fig.height=4, fig.width=4, cache=TRUE,fig.cap="Belted Kingfisher"}

#Megaceryle alcyon - belted kingfisher
mea<- dat%>%
  filter(species=="Megaceryle alcyon")%>%
  group_by(year)%>%
  mutate(date=as.Date(paste0(year,"-",month,"-",day)),
         j.day=julian(date,origin=as.Date(paste0(unique(year),"-01-01")))
  )%>%
  group_by(species,year,j.day,date)%>%
  summarise(day.tot=sum(individualCount,na.rm=T))%>%
  group_by(species,year)%>%
  mutate(prop=cumsum(day.tot/sum(day.tot,na.rm = T)))%>%
  filter(year>1999)
mea.pred <- mea%>%
  group_by(year)%>%
  summarize(
    pred=predict(nls(prop~SSlogis(j.day,Asym, xmid, scal)),newdata=data.frame(j.day=min(j.day):max(j.day))),#predict the logistic curve for each species
    j.day=min(j.day):max(j.day),
  )%>%
  left_join(mea%>%dplyr::select(j.day,date)) ## add date back to tibble

mea%>%
  ggplot(aes(j.day,prop))+geom_point(aes=0.3)+geom_line(data=mea.pred,aes(x=j.day,y=pred),col="steelblue",size=2)+facet_wrap(year~.)
```

```{r,"species ep", fig.height=4, fig.width=4, cache=TRUE,fig.cap="Eastern Phoebe"}
ep<- dat%>%
  filter(species=="Sayornis phoebe")%>%
  group_by(year)%>%
  mutate(date=as.Date(paste0(year,"-",month,"-",day)),
         j.day=julian(date,origin=as.Date(paste0(unique(year),"-01-01")))
  )%>%
  group_by(species,year,j.day,date)%>%
  summarise(day.tot=sum(individualCount,na.rm=T))%>%
  group_by(species,year)%>%
  mutate(prop=cumsum(day.tot/sum(day.tot,na.rm = T)))%>%
  filter(year>1999)
ep.pred <- ep%>%
  group_by(year)%>%
  summarize(
    pred=predict(nls(prop~SSlogis(j.day,Asym, xmid, scal)),newdata=data.frame(j.day=min(j.day):max(j.day))),#predict the logistic curve for each species
    j.day=min(j.day):max(j.day),
  )%>%
  left_join(ep%>%dplyr::select(j.day,date)) ## add date back to tibble

ep%>%
  ggplot(aes(j.day,prop))+geom_point(aes=0.3)+geom_line(data=ep.pred,aes(x=j.day,y=pred),col="navajowhite4",size=2)+facet_wrap(year~.)
```

```{r, "species PV", fig.height=4, fig.width=4, cache=TRUE,fig.cap="Philadelphia Vireo"}
pv<- dat%>%
  filter(species=="Vireo philadelphicus")%>%
  group_by(year)%>%
  mutate(date=as.Date(paste0(year,"-",month,"-",day)),
         j.day=julian(date,origin=as.Date(paste0(unique(year),"-01-01")))
  )%>%
  group_by(species,year,j.day,date)%>%
  summarise(day.tot=sum(individualCount,na.rm=T))%>%
  group_by(species,year)%>%
  mutate(prop=cumsum(day.tot/sum(day.tot,na.rm = T)))%>%
  filter(year>1999)

#Vireo philadelphicus - Philadelphia Vireo 
pv.forlog <- pv %>%  filter(year==2001 | year == 2002 | year == 2005|year ==2007|year==2011|year==2017|year==2018|year==2019)
pv.pred <- pv.forlog%>%
  group_by(year)%>%
  summarize(
    pred=predict(nls(prop~SSlogis(j.day,Asym, xmid, scal)),newdata=data.frame(j.day=min(j.day):max(j.day))),#predict the logistic curve for each species
    j.day=min(j.day):max(j.day),
  )%>%
  left_join(pv.forlog%>%dplyr::select(j.day,date)) ## add date back to tibble
pv.forlog%>%
  ggplot(aes(j.day,prop))+geom_point(aes=0.3)+geom_line(data=pv.pred,aes(x=j.day,y=pred),col="yellow4",size=2)+facet_wrap(year~.)
```

There are fewer years than the other species for the Philadelphia Vireo. This is due to the fact that there were too few occurrence data points for the omitted years to base a logistic curve off of. This is prevalent in a few other species as well. 

```{r, "species st", fig.height=4, fig.width=4, cache=TRUE,fig.cap="Swainson's Thrush"}
st <-  dat%>%
  filter(species=="Catharus ustulatus")%>%
  group_by(year)%>%
  mutate(date=as.Date(paste0(year,"-",month,"-",day)),
         j.day=julian(date,origin=as.Date(paste0(unique(year),"-01-01")))
  )%>%
  group_by(species,year,j.day,date)%>%
  summarise(day.tot=sum(individualCount,na.rm=T))%>%
  group_by(species,year)%>%
  mutate(prop=cumsum(day.tot/sum(day.tot,na.rm = T)))%>%
  filter(year>1999)
st.pred <- st%>%
  group_by(year)%>%
  summarize(
    pred=predict(nls(prop~SSlogis(j.day,Asym, xmid, scal)),newdata=data.frame(j.day=min(j.day):max(j.day))),#predict the logistic curve for each species
    j.day=min(j.day):max(j.day),
  )%>%
  left_join(st%>%dplyr::select(j.day,date)) ## add date back to tibble
st%>%
  ggplot(aes(j.day,prop))+geom_point(aes=0.3)+geom_line(data=st.pred,aes(x=j.day,y=pred),col="khaki4",size=2)+facet_wrap(year~.)
```

```{r, "speciesbtw", fig.height=4, fig.width=4, cache=TRUE,fig.cap="Black-Throated Blue Warbler"}
#Setophaga caerulescens - Black-throated Blue Warbler 
btw <-  dat%>%
  filter(species=="Setophaga caerulescens")%>%
  group_by(year)%>%
  mutate(date=as.Date(paste0(year,"-",month,"-",day)),
         j.day=julian(date,origin=as.Date(paste0(unique(year),"-01-01")))
  )%>%
  group_by(species,year,j.day,date)%>%
  summarise(day.tot=sum(individualCount,na.rm=T))%>%
  group_by(species,year)%>%
  mutate(prop=cumsum(day.tot/sum(day.tot,na.rm = T)))%>%
  filter(year>1999)
btw.pred <- btw%>%
  group_by(year)%>%
  summarize(
    pred=predict(nls(prop~SSlogis(j.day,Asym, xmid, scal)),newdata=data.frame(j.day=min(j.day):max(j.day))),#predict the logistic curve for each species
    j.day=min(j.day):max(j.day),
  )%>%
  left_join(btw%>%dplyr::select(j.day,date)) ## add date back to tibble
btw%>%
  ggplot(aes(j.day,prop))+geom_point(aes=0.3)+geom_line(data=btw.pred,aes(x=j.day,y=pred),col="royalblue3",size=2)+facet_wrap(year~.)
```

```{r, "species yw", fig.height=4, fig.width=4, cache=TRUE,fig.cap="Yellow-Throated Warbler"}
#Setophaga dominica - Yellow-throated  Warbler 
yw <-  dat%>%
  filter(species=="Setophaga dominica")%>%
  group_by(year)%>%
  mutate(date=as.Date(paste0(year,"-",month,"-",day)),
         j.day=julian(date,origin=as.Date(paste0(unique(year),"-01-01")))
  )%>%
  group_by(species,year,j.day,date)%>%
  summarise(day.tot=sum(individualCount,na.rm=T))%>%
  group_by(species,year)%>%
  mutate(prop=cumsum(day.tot/sum(day.tot,na.rm = T)))%>%
  filter(year>1999)
yw.forlog <- yw%>%filter(year==2004 |year==2009 |year== 2011|year==2013| year==2014| year==2016| year==2017| year==2019)
yw.pred <- yw.forlog%>%
  group_by(year)%>%
  summarize(
    pred=predict(nls(prop~SSlogis(j.day,Asym, xmid, scal)),newdata=data.frame(j.day=min(j.day):max(j.day))),#predict the logistic curve for each species
    j.day=min(j.day):max(j.day),
  )%>%
  left_join(yw.forlog%>%dplyr::select(j.day,date)) ## add date back to tibble
yw.forlog%>%
  ggplot(aes(j.day,prop))+geom_point(aes=0.3)+geom_line(data=yw.pred,aes(x=j.day,y=pred),col="gold1",size=2)+facet_wrap(year~.)
```

From all of the species logistic models, it appears each species arrives at staggered points in the year. The next analysis will determine if these arrival dates fluctuate year to year, or if they are relatively consistent.

```{r fig.height=4, fig.width=4, cache=TRUE, "day of the year .25 of the population arrives, varying by year", fig.cap="Belted Kingfisher"}
#Megaceryle alcyon - belted kingfisher
mea.arrive.date <-mea.pred%>%
  group_by(year)%>%
  filter(j.day==j.day[which.min(abs(pred-0.25))])

mea.arrive.date%>%
  ggplot(aes(year,j.day))+geom_point()
```
```{r fig.height=4, fig.width=4, cache=TRUE, "day of the year .25 of the population arrives, varying 1 ", fig.cap="Eastern Phoebe"}
#Sayornis phoebe - Eastern Phoebe
ep.arrive.date <-ep.pred%>%
  group_by(year)%>%
  filter(j.day==j.day[which.min(abs(pred-0.25))])

ep.arrive.date%>%
  ggplot(aes(year,j.day))+geom_point()
```
```{r fig.height=4, fig.width=4, cache=TRUE, "day of the year .25 of the population arrives, varyin 2", fig.cap="Philadelphia Vireo"}
#Vireo philadelphicus - Philadelphia Vireo 
pv.arrive.date <-pv.pred%>%
  group_by(year)%>%
  filter(j.day==j.day[which.min(abs(pred-0.25))])

pv.arrive.date%>%
  ggplot(aes(year,j.day))+geom_point()
```
```{r fig.height=4, fig.width=4, cache=TRUE, "day of the year .25 of the population arrives, varying by year3", fig.cap="Swainson's Thrush"}
#Catharus ustulatus - Swainson's Thrush 
st.arrive.date <-st.pred%>%
  group_by(year)%>%
  filter(j.day==j.day[which.min(abs(pred-0.25))])

st.arrive.date%>%
  ggplot(aes(year,j.day))+geom_point()
```
```{r fig.height=4, fig.width=4, cache=TRUE, "day of the year .25 of the population arrives, varying by year4", fig.cap="Black-Throated Blue Warbler"}
#Setophaga caerulescens - Black-throated Blue Warbler 
btw.arrive.date <-btw.pred%>%
  group_by(year)%>%
  filter(j.day==j.day[which.min(abs(pred-0.25))])

btw.arrive.date%>%
  ggplot(aes(year,j.day))+geom_point()
```
```{r,fig.height=4, fig.width=4,"what day of the year .25 of the population arrives, varying by year5", fig.cap="Yellow-Throated Warbler",cache=TRUE}
#Setophaga dominica - Yellow-throated  Warbler 
yw.arrive.date <-yw.pred%>%
  group_by(year)%>%
  filter(j.day==j.day[which.min(abs(pred-0.25))])

yw.arrive.date%>%
  ggplot(aes(year,j.day))+geom_point()
```

These results insinuate that there is relative inconsistency among arrival date from year to year, for each species. We will conduct further analysis with weather data from the migration path to see if weather has an effect on these species' arrival dates on an annual basis. 

```{r,"making a tibble with weather data"}
weather.d <- weather.d%>%
  mutate(year=as.integer(str_sub(date,1,4)), #add year
         date=as.Date(date))%>%
  group_by(year)%>% #group by year so we can compute julian day
  mutate(j.day=julian(date,origin=as.Date(paste0(unique(year),"-01-01"))), #add julian day
         date2=date,
         wdir.rad=(180-abs(wdf2-180))*pi/180, #radians so we can use a trig function to compute wind vector, scale degrees first to 180 scale to 2x pi and subtract from 180 (wind comes out of a direction)
         wvec=cos(wdir.rad)*-1*awnd # we want a negative value for positive value for 2x pi
  )%>% #store day in new column
  dplyr::select(id,year,date2,j.day,tmin,tmax,wvec)%>% #select the rows we need
  left_join(sta.d%>%select(id,name,migr.day))%>% #add the station id info (ie. name)
  mutate(j.day=j.day+migr.day)#make j.day ahead of BOS according to the migration days away so we can join weather along path
```
```{r,"combining weather data and species arrival da"}
#Megaceryle alcyon - belted kingfisher
mea.arr.weath <- mea.arrive.date%>%
  left_join(weather.d)%>%
  left_join(mea%>%dplyr::select(year,date,j.day))

#eastern phoebe 
ep.arr.weath <- ep.arrive.date%>%
  left_join(weather.d)%>%
  left_join(ep%>%dplyr::select(year,date,j.day))

#Vireo philadelphicus - Philadelphia Vireo 
pv.arr.weath <- pv.arrive.date%>%
  left_join(weather.d)%>%
  left_join(pv%>%dplyr::select(year,date,j.day))

#Catharus ustulatus - Swainson's Thrush 
st.arr.weath <- st.arrive.date%>%
  left_join(weather.d)%>%
  left_join(st%>%dplyr::select(year,date,j.day))

#Setophaga caerulescens - Black-throated Blue Warbler 
btw.arr.weath <- btw.arrive.date%>%
  left_join(weather.d)%>%
  left_join(btw%>%dplyr::select(year,date,j.day))

#Setophaga dominica - Yellow-throated  Warbler
yw.arr.weath <- yw.arrive.date%>%
  left_join(weather.d)%>%
  left_join(yw%>%dplyr::select(year,date,j.day))
```
```{r,"finding mean of weather variables, over time span of migration through stations"}
weather.wk <-weather.d %>% 
  group_by(year,name) %>% 
  mutate(wk.tmin = frollmean(tmin, n=14,align="right"),
         wk.tmax = frollmean(tmax, n=14,align="right"),
         wk.wvec = frollmean(wvec, n=14,align="right")
  )%>%
  dplyr::select(j.day,date2,name,wk.tmin,wk.tmax,wk.wvec)
```
```{r,"combining mean weather data with species arrival data"}
mea.arr.weath2 <- mea.arrive.date%>%
  left_join(weather.wk)
ep.arr.weath2 <- ep.arrive.date%>%
  left_join(weather.wk)
pv.arr.weath2 <- pv.arrive.date%>%
  left_join(weather.wk)
st.arr.weath2 <- st.arrive.date%>%
  left_join(weather.wk)
btw.arr.weath2 <- btw.arrive.date%>%
  left_join(weather.wk)
yw.arr.weath2 <- yw.arrive.date%>%
  left_join(weather.wk)
```

### Linear Models

```{r,"linear models,non-averaged weather data"}
mea.lmer <- lmer(j.day~tmin*tmax*wvec+(1|name),mea.arr.weath,na.action = "na.fail")
anova.mea <- Anova(mea.lmer)
kable(anova.mea,caption= "Belted Kingfisher")

ep.lmer <- lmer(j.day~tmin*tmax*wvec+(1|name),ep.arr.weath,na.action = "na.fail")
anova.ep <- Anova(ep.lmer)
kable(anova.ep,caption = "Eastern Phoebe")

pv.lmer <- lmer(j.day~tmin*tmax*wvec+(1|name),pv.arr.weath,na.action = "na.fail")
anova.pv <- Anova(pv.lmer)
kable(anova.pv,caption="Philadelphia Vireo")

st.lmer <- lmer(j.day~tmin*tmax*wvec+(1|name),st.arr.weath,na.action = "na.fail")
anova.st <- Anova(st.lmer)
kable(anova.st,caption="Swainson's Thrush")

btw.lmer <- lmer(j.day~tmin*tmax*wvec+(1|name),btw.arr.weath,na.action = "na.fail")
anova.btw <- Anova(btw.lmer)
kable(anova.btw,caption="Black Throated Blue Warbler")

yw.lmer <- lmer(j.day~tmin*tmax*wvec+(1|name),yw.arr.weath,na.action = "na.fail")
anova.yw <- Anova(yw.lmer)
kable(anova.yw,caption="Yellow-Throated Warbler")
```

Due to the nature of our complex data set, we performed linear mixed-effect modeling, as there are some variables (such as the weather variables of temperature and the wind vector) that are fixed, meaning they contribute to the slope and intercept of model but do not change among all of the samples we have; i.e. the weather is the same for each bird species. There are random effect variables, like the location of the station and the species, which we also have to take into consideration when we create models to better understand trends in our data. The first linear model assessed used weather data from a single-day report from the weather station to assess the relationship of those fixed weather variables with the arrival date (j.day), with a random effect of the station location. The weather variables assessed are maximum temperature of the area (tmax), minimum temperature of the area (tmin), and then a wind vector value. We calculated using average wind velocity in meters per second and the direction of the fastest 2-minute wind.  We performed ANOVA tests to see how the linear model created fits the data. The ANOVA test results indicate using only single-day weather data at each location doesn’t reveal any significant relationships between weather and arrival day. 
```{r,"linear models, mean-weather  2 weeks"}
mea.lmer2 <- lmer(j.day~wk.tmin*wk.tmax*wk.wvec+(1|name),mea.arr.weath2,na.action = "na.fail")
anova.mea2 <- Anova(mea.lmer2)
kable(anova.mea2,caption = "Belted Kingfisher")

ep.lmer2 <- lmer(j.day~wk.tmin*wk.tmax*wk.wvec+(1|name),ep.arr.weath2,na.action = "na.fail")
anova.ep2 <- Anova(ep.lmer2)
kable(anova.ep2,caption="Eastern Phoebe")

pv.lmer2 <- lmer(j.day~wk.tmin*wk.tmax*wk.wvec+(1|name),pv.arr.weath2,na.action = "na.fail")
anova.pv2 <- Anova(pv.lmer2)
kable(anova.pv2,caption="Philadelphia Vireo")

st.lmer2 <- lmer(j.day~wk.tmin*wk.tmax*wk.wvec+(1|name),st.arr.weath2,na.action = "na.fail")
anova.st2 <- Anova(st.lmer2)
kable(anova.st2,caption="Swainson's Thrush")

btw.lmer2 <- lmer(j.day~wk.tmin*wk.tmax*wk.wvec+(1|name),btw.arr.weath2,na.action = "na.fail")
anova.btw2 <- Anova(btw.lmer2)
kable(anova.btw2,caption="Black-Throated Blue Warbler")

yw.lmer2 <- lmer(j.day~wk.tmin*wk.tmax*wk.wvec+(1|name),yw.arr.weath2,na.action = "na.fail")
anova.yw2 <- Anova(yw.lmer2)
kable(anova.yw2,caption="Yellow-Throated Warbler")
```

In this second set of linear models, we had the same model (with fixed weather variable effects and random effect of station location), but instead of using weather data from single-days in time, we calculated averages of the weather data variables over the span of 2 weeks, to account for any birds that do not all arrive / leave at the same time. With these weather averages, there do appear to be significant fixed effects of weather, in many of the species -- all three of the variables seem to have a relationship with the arrival date per species. 

### Secondary Model with Single-Day Weather data variables
```{r,"secondary models"}
mea.arr.aic1 <- dredge(mea.lmer,fixed = c("tmin","tmax","wvec"),)
kable(mea.arr.aic1[1:4,],caption = "Fit values for nested models of the most complicated lme model for Belted Kingfisher")

ep.arr.aic1 <- dredge(ep.lmer,fixed = c("tmin","tmax","wvec"),)
kable(ep.arr.aic1[1:4,],caption = "Fit values for nested models of the most complicated lme model for Eastern Phoebe")

pv.arr.aic1 <- dredge(pv.lmer,fixed = c("tmin","tmax","wvec"),)
kable(pv.arr.aic1[1:4,],caption = "Fit values for nested models of the most complicated lme model for Philadelphia Vireo")

st.arr.aic1 <- dredge(st.lmer,fixed = c("tmin","tmax","wvec"),)
kable(st.arr.aic1[1:4,],caption = "Fit values for nested models of the most complicated lme model for Swainson's Thrush")

btw.arr.aic1 <- dredge(btw.lmer,fixed = c("tmin","tmax","wvec"),)
kable(btw.arr.aic1[1:4,],caption = "Fit values for nested models of the most complicated lme model for Black-Throated Blue Warbler")

yw.arr.aic1 <- dredge(yw.lmer,fixed = c("tmin","tmax","wvec"),)
kable(yw.arr.aic1[1:4,],caption = "Fit values for nested models of the most complicated lme model for Yellow-Throated Warbler")
```

We utilized another method to find a best fit model, where we used the dredge function to assess only the three weather variables: tmin, tmax, and wvec, and their varying combinations of interactions. The data tables above indicate the best fitting models with the 3 weather variables as fixed effects. From these tables, we can surmise that the best fitting models, for each species, contain all three variables to have an effect on arrival date. 

### Secondary Model with Average weather data variables 

```{r,"secondary models2"}
mea.arr.aic <- dredge(mea.lmer2,fixed = c("wk.tmin","wk.tmax","wk.wvec"),)
kable(mea.arr.aic[1:4,],caption = "Fit values for nested models of the most complicated lme model for Belted Kingfisher")

ep.arr.aic <- dredge(ep.lmer2,fixed = c("wk.tmin","wk.tmax","wk.wvec"),)
kable(ep.arr.aic[1:4,],caption = "Fit values for nested models of the most complicated lme model for Eastern Phoebe")

pv.arr.aic <- dredge(pv.lmer2,fixed = c("wk.tmin","wk.tmax","wk.wvec"),)
kable(pv.arr.aic[1:4,],caption = "Fit values for nested models of the most complicated lme model for Philadelphia Vireo")

st.arr.aic <- dredge(st.lmer2,fixed = c("wk.tmin","wk.tmax","wk.wvec"),)
kable(st.arr.aic[1:4,],caption = "Fit values for nested models of the most complicated lme model for Swainson's Thrush")

btw.arr.aic <- dredge(btw.lmer2,fixed = c("wk.tmin","wk.tmax","wk.wvec"),)
kable(btw.arr.aic[1:4,],caption = "Fit values for nested models of the most complicated lme model for Black-Throated Blue Warbler")

yw.arr.aic <- dredge(yw.lmer2,fixed = c("wk.tmin","wk.tmax","wk.wvec"),)
kable(yw.arr.aic[1:4,],caption = "Fit values for nested models of the most complicated lme model for Yellow-Throated Warbler")
```

Similar to the secondary model utilizing single-day weather data variables, this model represents the best-fitting models using the 2-week average analysis weather variables. The data tables above indicate the best fitting models with the 3 weather variables as fixed effects. From these tables, we can surmise that the best fitting models, for each species, reflect all three variables having an effect on arrival date. 

### Best fitting models


```{r,"best fitting models -- single day data"}
best.mea.lmer1 <-  lmer(j.day~tmin+tmax+wvec+(1|name),mea.arr.weath,na.action = "na.fail")
bestmea1 <- Anova(best.mea.lmer1)
kable(bestmea1,caption="Belted Kingfisher")
best.ep.lmer1 <-  lmer(j.day~tmin+tmax+wvec+(1|name),ep.arr.weath,na.action = "na.fail")
bestep1 <- Anova(best.ep.lmer1)
kable(bestep1,caption="Eastern Phoebe")
best.pv.lmer1 <-  lmer(j.day~tmin+tmax+wvec+(1|name),pv.arr.weath,na.action = "na.fail")
bestpv1 <- Anova(best.pv.lmer1)
kable(bestpv1,caption="Philadelphia Vireo")
best.st.lmer1 <-  lmer(j.day~tmin+tmax+wvec+(1|name),st.arr.weath,na.action = "na.fail")
bestst1 <- Anova(best.st.lmer1)
kable(bestst1,caption="Swainson's Thrush")
best.btw.lmer1 <-  lmer(j.day~tmin+tmax+wvec+(1|name),btw.arr.weath,na.action = "na.fail")
bestbtw1 <- Anova(best.btw.lmer1)
kable(bestbtw1,caption="Black-Throated Blue Warbler")
best.yw.lmer1 <-  lmer(j.day~tmin+tmax+wvec+(1|name),yw.arr.weath,na.action = "na.fail")
bestyw1 <- Anova(best.yw.lmer1)
kable(bestyw1,caption="Yellow-Throated Warbler")
```
### Best fitting models -- average weather data

```{r,"best fitting models average weather data"}
best.mea.lmer <-  lmer(j.day~wk.tmin+wk.tmax+wk.wvec+(1|name),mea.arr.weath2,na.action = "na.fail")
bestmea <- Anova(best.mea.lmer)
kable(bestmea,caption="Belted Kingfisher")
best.ep.lmer <-  lmer(j.day~wk.tmin+wk.tmax+wk.wvec+(1|name),ep.arr.weath2,na.action = "na.fail")
bestep <- Anova(best.ep.lmer)
kable(bestep,caption="Eastern Phoebe")
best.pv.lmer <-  lmer(j.day~wk.tmin+wk.tmax+wk.wvec+(1|name),pv.arr.weath2,na.action = "na.fail")
bestpv <- Anova(best.pv.lmer)
kable(bestpv,caption="Philadelphia Vireo")
best.st.lmer <-  lmer(j.day~wk.tmin+wk.tmax+wk.wvec+(1|name),st.arr.weath2,na.action = "na.fail")
bestst <- Anova(best.st.lmer)
kable(bestst,caption="Swainson's Thrush")
best.btw.lmer <-  lmer(j.day~wk.tmin+wk.tmax+wk.wvec+(1|name),btw.arr.weath2,na.action = "na.fail")
bestbtw <- Anova(best.btw.lmer)
kable(bestbtw,caption="Black-Throated Blue Warbler")
best.yw.lmer <-  lmer(j.day~wk.tmin+wk.tmax+wk.wvec+(1|name),yw.arr.weath2,na.action = "na.fail")
bestyw <- Anova(best.yw.lmer)
kable(bestyw,caption="Yellow-Throated Warbler")
```

These tables dispaly the ANOVA values for the best-fit models found using only the 3 weather variables as fixed effects (secondary model). There appears to be some significance for these weather variables having a fixed effect on the arrival date of the species. 

### Additional analysis: Looking at species as a random effect variable 

There is a lot more analysis that could be conducted with this data set we have, but specifically, we thought it might be interesting to touch upon the effect that species has on arrival date. Similar to the method we conducted above per species, using all of the data, we created a model that used both location of station and species as random effects to assess the relationship between arrival date and weather variables. 

```{r,"additional analysis",cache=TRUE}
#establishing arrival date for estimate of all species logistic model 
all.dat.pred <- all.dat%>%
  group_by(year)%>%
  summarize(
    pred=predict(nls(prop~SSlogis(j.day,Asym, xmid, scal)),newdata=data.frame(j.day=min(j.day):max(j.day))),#predict the logistic curve for each species
    j.day=min(j.day):max(j.day),
  )%>%
  left_join(all.dat%>%dplyr::select(j.day,date)) ## add date back to tibble
alldat.arrive.date <-all.dat.pred%>%
  group_by(year)%>%
  filter(j.day==j.day[which.min(abs(pred-0.25))])
#combining with weater means (staggered bc they're all coming at different times)
alldat.arr.weath <- alldat.arrive.date%>%
  left_join(weather.d)%>%
  left_join(all.dat%>%dplyr::select(year,date,j.day))
alldat.arr.weath2 <- alldat.arrive.date%>%
  left_join(weather.wk)
#linear model 1 
alldat.lmer2 <- lmer(j.day~wk.tmin*wk.tmax*wk.wvec+(1|name)+(1|species),alldat.arr.weath2,na.action = "na.fail")
anova.all <- Anova(alldat.lmer2)
kable(anova.all,caption="ANOVA values for linear model of all 6 species")
#secondary model 
alldat.arr.aic <- dredge(alldat.lmer2,fixed = c("wk.tmin","wk.tmax","wk.wvec"),)
kable(pv.arr.aic[1:4,],caption = "Fit values for nested models of the most complicated lme model for all species")
#all data best model 
best.alldat.lmer <-  lmer(j.day~wk.tmin+wk.tmax+wk.wvec+(1|name)+(1|species),alldat.arr.weath2,na.action = "na.fail")
anova.bestmodel <- Anova(best.alldat.lmer)
kable(anova.bestmodel,caption = "ANOVA values for best fit linear model of all species")
```

The significant values in the ANOVA table for the various linear models indicate that there most likely exists a relationship between the weather and arrival date of all species -- even with the random effects of location and species.  


## Discussion 

Following analysis, it is possible to assess how arrival time of six trans-Gulf migrant (TGMs) species (Belted Kingfisher, Eastern Phoebe, Philadelphia Vireo, Swainson's Thrush, Black-Throated Blue Warbler, and Yellow-Throated Warbler) is affected by meteorological conditions, such as temperature and wind and how this variation in arrival time may affect population status. Since variation of arrival time can impact mating success, available food supply, and more, this topic is pertinent to the survival of trans-Gulf migrants as climate change further disrupts typical weather patterns. As demonstrated in our results, for all species studied, there are inconsistencies in the arrival dates of TGMs between 2000 and 2019. In the linear models that analyzed the averages of the weather data over the course of two weeks, there are significant effects of weather, in all three variables of minimum temperature, maximum temperature, and wind velocity, on arrival day of the chosen TGMs. The models of best fit for each species analyzed contain all three of these weather variables as having an effect on arrival date. Overall, this analysis demonstrated that weather conditions do have an effect on Julian day of various TGMs. However, we have yet to fully establish in which direction population status is affected by meteorological conditions. In relation to previous literature, our results confirm the general consensus. @schaub2004departure studied the departure of migrating European robins and how it was affected by rain and wind velocity. Despite the difference in species and location between our analysis and this study, similar conclusions were drawn. There was clear evidence that wind velocity had an effect on arrival times, and subsequently departure times, of the European robins. Rain also significantly affected migration both spatially and temporally. However, @schaub2004departure also emphasizes that it is not always clear in what direction these meteorological factors will impact migration patterns and arrival days. In terms of temperature, @hurlbert2012spatiotemporal utilized citizen science data from eBird as well. In a similar vain to our results, they found that for all species in all geographic regions investigated, arrival date shifted earlier 0.8-6 days for every one degree Celsius increase in temperature. As our results confirm the results of previous studies, the major takeaway is that there is a link between meteorological factors, like temperature and wind velocity, and the Julian day of TGMs. Migratory birds will likely suffer due to mating patterns and food availability being routed in weather patterns. However, this effect on population status is not well understood yet. Other rhythms of development rooted in migratory patterns are gonad growth and rapid fat accumulation, according to @hurlbert2012spatiotemporal, which has the potential to lead to grave physiological consequences as the severity of this issue increases. It is imperative that further studies continue to be undertaken to gain a more comprehensive understanding of how meteorological changes affect the arrival time of passerine birds along their migration route and what implications this has for population levels. 

## Author Contributions 
Michael Britt - Bibliography, Discussion

Phoebe Chandler - Methods

Ellie Falanga - Results and Code

Erin Kiley - Introduction

## References 

